### 2016-02-14 text/Chapter05.tex
Add implicit parameters and generalization (Chapter 5)

We first consider dataflow computations. As discussed earlier in the
semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
values. When reducing translated programs, the comonadic values will not be directly manipulated by the
user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
operations are the comonadically-inspired ones defined earlier.

The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
and a matching kind of expressions. We add a language primitive modelling the semantics of the
\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
that the number of elements in the stream matches the number in the coeffect.

The additional reduction rules mirror the semantics that we discussed earlier when talking about
indexed dataflow comonad.

@@

\begin{equation*}
\begin{array}{rcl}
v &=& \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
e &=& \ldots \sep \ident{Df}\langle e_0, \ldots, e_n\rangle \sep \ident{prev}_{\cclrd{n}}~e\\
C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_
\end{array}  
\end{equation*}

@@

\begin{theorem}[Well-typedness of the \kvd{prev} translation]
\label{thm:semantics-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
and Figure~\ref{fig:semantics-ext-df} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation.

@@

@@

::::TODO::::

We first consider dataflow computations. As discussed earlier in the
semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
values. When reducing translated programs, the comonadic values will not be directly manipulated by the
user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
operations are the comonadically-inspired ones defined earlier.

The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
and a matching kind of expressions. We add a language primitive modelling the semantics of the
\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
that the number of elements in the stream matches the number in the coeffect.

The additional reduction rules mirror the semantics that we discussed earlier when talking about
indexed dataflow comonad.

\paragraph{Properties.}

@@

\begin{equation*}
\begin{array}{rcl}
v &=& \ldots \sep \ident{Impl}(v, \langle \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n\rangle) \\
e &=& \ldots \sep \ident{Impl}(e, \langle\ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n\rangle) \\
  && \hspace{1.4em} \sep \ident{lookup}_{\cclrd{\ident{?p}}}~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~e_1~e_2\\
C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_
\end{array}  
\end{equation*}

\begin{equation*}
\tyrule{impl}
  { \Gamma \vdash e:\tau & \forall i\in\{ 1 \ldots n \}. ~ \Gamma \vdash e_i : \ident{num}}
  { \Gamma \vdash \ident{Impl}(e, \langle \ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n \rangle) : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \}}}{\tau} }
\end{equation*}
\begin{equation*}
\tyrule{lookup}
  { \Gamma \vdash e : \ctyp{\cclrd{ \{ \ldots, \ident{?p}, \ldots \} }}{\tau} }
  { \Gamma \vdash \ident{lookup}_{\cclrd{\ident{?p}}}~e : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{letimpl}
  { \Gamma \vdash e_1 : \ident{num} &
    \Gamma \vdash e_2 : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}{\tau} }
  { \Gamma \vdash \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}~e_1~e_2 : 
      \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n, \ident{?p} \} }}{\tau} }
\end{equation*}

\begin{equation*}
\semdef
  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
  {\lambda ctx.\ident{lookup}_{\cclrd{\ident{?p}}}~\ctx }
\end{equation*}

\begin{equation*}
\semdefff
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\hspace{-1em}\begin{array}{l}  
      \coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} }\\[-0.3em]
        \quad \vdash \kvd{let}~\ident{?p}=e_1\\[-0.3em]
        \quyad\qquad \kvd{in}~e_2 : \tau_2 \\[-0.3em]
        ~ \\[-0.3em]
   \end{array}}
  {f}
  {g}
  {\hspace{-1em}\begin{array}{l}  
  \lambda\ctx.~~\\[-0.3em]
    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cclrd{\cup}\; \cclrd{(s\setminus{\ident{?p}})}}~\ident{dup}~\ctx\\[-0.3em]
    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{(s\setminus{\ident{?p}})}}~\ctx_0\\[-0.3em]
    \quad g~(\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{(s\setminus{\ident{?p}})}}~(f~\ctx_1)~\ctx_2)
  \end{array} }   
\end{equation*}

::: TODO ::: Everything except for counit

\begin{equation*}
\begin{array}{rl}
{\footnotesize(\emph{counit})} & \begin{array}{l}
\ident{counit}_{\cclrd{ \emptyset }}~\ident{Impl} (v, \langle \ldots \rangle)   \rightarrow v
\end{array}
\\[0.5em]
{\footnotesize(\emph{cobind})} & \begin{array}{l}
\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
\\[-0.25em]
\quad (\ident{Df}\langle f (\ident{Df}\langle v_0, \ldots, v_m \rangle), \ldots, f (\ident{Df}\langle v_{n}, \ldots, v_{m+n} \rangle) \rangle)
\end{array}
\\[0.5em]
{\footnotesize(\emph{merge})} & \begin{array}{l}
\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightarrow
\\[-0.25em]
\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, v'_{\mathit{min}(m,n)}) \rangle)
\end{array}
\\[0.5em]
{\footnotesize(\emph{split})} & \begin{array}{l}
\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightarrow
\\[-0.25em]
\quad \ident{Df}\langle v_0, \ldots, v_m \rangle, (\ident{Df}\langle b_0, \ldots, b_n\rangle
\end{array}
\\[0.5em]
{\footnotesize(\emph{prev})} & \begin{array}{l}
\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightarrow
\\[-0.25em]
\quad \ident{Df}\langle v_0, \ldots, v_n \rangle
\end{array}
\end{array}
\end{equation*}

\caption{Additional constructs embedding implicit parameters into the language}
\label{fig:semantics-ext-df}
\end{figure}
\newpage

@@

\section{Generalising}
\label{sec:semantics-generalising}
... works for any good comonad...

\newpage
\newpage

>>>>

We first consider dataflow computations. As discussed earlier in the
semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
values. When reducing translated programs, the comonadic values will not be directly manipulated by the
user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
operations are the comonadically-inspired ones defined earlier, together with an additional 
\emph{domain-specific} operation that models the \kvd{prev} construct.

The Figure~\ref{fig:semantics-ext-df} extensionds the target language with syntax, typing rules,
additional translation rule and reductions for modelling dataflow computations. We introduce a 
new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
and a matching kind of expressions. We specify how the \kvd{prev} keyword is translated into a
$\ident{prev}_\cclrd{r}$ operation of the target language nd we also add a typing rule
(\emph{df}) that checks the types of the elements of the stream and also guarantees 
that the number of elements in the stream matches the number in the coeffect.
The additional reduction rules mirror the semantics that we discussed earlier when talking about
indexed dataflow comonad.

@@

\begin{equation*}
\begin{array}{rcl}
v &=& \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
e &=& \ldots \sep \ident{Df}\langle e_0, \ldots, e_n\rangle \sep \ident{prev}_{\cclrd{n}}~e\\
C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_ \sep \ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle
\end{array}  
\end{equation*}

@@

\begin{theorem}[Well-typedness of the \kvd{prev} translation]
\label{thm:semantics-df-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
and Figure~\ref{fig:semantics-ext-df} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation.

@@

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{df}): $e = \ident{Df}\langle e_0, \ldots, e_n\rangle$. If $e_i$ is not a value
  then reduce using (\emph{ctx}) with context $\ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle$.
  Otherwise, $e_0, \ldots, e_n$ are values and so $\ident{Df}\langle e_0, \ldots, e_n\rangle$
  is also a value.

@@

As discussed earlier (Example~\ref{thm:semantics-indexed-prod}), implicit parameters can be modelled 
by an indexed product comonad, which annotates a value with additional context -- in our case, a 
mapping from implicit parameter names to their values. In this section, we embed this model into
the target language.

As with dataflow computations, we take the core functional subset (Figure~\ref{fig:semantics-target}) 
with comonadically-inspired extensions (Figure~\ref{fig:semantics-ext}) and we specify a new kind
of values of type $\ctyp{\cclrd{r}}\tau$ and domain-specific reduction rules that specify how
the operations propagate and access the context containing implicit parameter bindings.
Again, the $\ctyp{\cclrd{r}}\tau$ values can be seen as an \emph{abstract data type}, which are 
never manipulated directly.

\paragraph{Domain-specific extensions.}

The Figure~\ref{fig:semantics-ext-impl} shows extensions to the target language for modelling
implicit parameters. A comonadic value with a coeffect $\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}$
is modelled by a new kind of value written as 
$\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots \ident{?p}_n \mapsto v_n \})$ which contains
a value $v$ together with implicit parameter assignments for all the parameters specified in the
coeffect. We add a corresponding kind of expression with its typing rule (\emph{impl}).

There are also two domain-specific operations for working with implicit parameters. The
$\ident{lookup}_{\cclrd{\ident{?p}}}$ operation reads a value of an implicit parameter and
the $\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}$ operation adds a mapping assigning a value
to an implicit parameter \ident{?p}. The typing rule (\emph{lookup}) specifies that the accessed
parameter need to be a part of the context and the rule (\emph{letimpl}) specifies that the
\ident{letimpl} operation extends the context with a new implicit parameter binding.

The new translation rules specify how implicit parameter access, written as \ident{?p}, and
implicit parameter binding, written as $\kvd{let}~\ident{?p}~=e_1~\kvd{in}~e_2$ are translated to the
target language. The first one is straightforward. The binding is similar to the translation for
function application -- we split the context, evaluate $e_1$ using the first part of the context
$\ctx_1$ and then add the new binding to the remaining context $\ctx_2$.

Finally, Figure~\ref{fig:semantics-ext-impl} also defines the reduction rules. The (\emph{lookup})
rule accesses an implicit parameter and (\emph{letimpl}) adds a new binding. The reduction rules 
for comonadically-inspired primitives closely model the product comonad discussed in 
Example~\ref{thm:semantics-indexed-prod}. Reductions for (\emph{cobind}) and (\emph{split}) 
restrict the set of available implicit parameters according to the annotations and 
(\emph{merge}) combines them, preferring the values from the call-site.

For the semantics of implicit parameter programs that we consider, the preference of call-site
bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing 
derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always
split implicit parameters into \emph{disjoint sets}.

@@

\begin{equation*}
\begin{array}{rcl}
v &=& \ldots \sep \ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n\}) \\
e &=& \ldots \sep \ident{Impl}(e, \{\ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n\}) \\
  && \hspace{1.4em} \sep \ident{lookup}_{\cclrd{\ident{?p}}}~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~e_1~e_2\\
C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_ \\
  && \hspace{1.4em} \sep \ident{Impl}(\_, \{ \ident{?p}_1\mapsto e_1, .., \ident{?p}_n\mapsto e_n\})\\
  && \hspace{1.4em} \sep \ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, .., \ident{?p}_{i-1} \mapsto v_{i-1}, \ident{?p}_i \mapsto \_, \ident{?p}_{i+1} \mapsto v_{i+1}, .. \ident{?p}_n \mapsto e_n\})
\end{array}  
\end{equation*}

\vspace{1em}
\paragraph{Typing rules}

\begin{equation*}
  \hspace{-2em}
\tyrule{impl}
  { \Gamma \vdash e:\tau & \forall i\in\{ 1 \ldots n \}. ~ \Gamma \vdash e_i : \ident{num}}
  { \Gamma \vdash \ident{Impl}(e, \{ \ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n \}) : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \}}}{\tau} }
\end{equation*}
\begin{equation*}
  \hspace{-2em}
\tyrule{lookup}
  { \Gamma \vdash e : \ctyp{\cclrd{ \{\ident{?p}\} }}{\tau} }
  { \Gamma \vdash \ident{lookup}_{\cclrd{\ident{?p}}}~e : \ident{num} }
\end{equation*}
\begin{equation*}
  \hspace{-2em}
\tyrule{letimpl}
  { \Gamma \vdash e_1 : \ident{num} &
    \Gamma \vdash e_2 : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}{\tau} }
  { \Gamma \vdash \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}~e_1~e_2 : 
      \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n, \ident{?p} \} }}{\tau} }
\end{equation*}

\vspace{1em}
\paragraph{Translation}

\begin{equation*}
\begin{array}{rl}  
(\footnotesize{\emph{lookup}}) &
\semdef
  { \coctx{\Gamma}{\cclrd{ \{\ident{?p}\} }} \vdash \ident{?p} : \ident{num} }
  {\lambda ctx.\ident{lookup}_{\cclrd{\ident{?p}}}~\ctx }
\\[2em]
\begin{array}{l}(\footnotesize{\emph{letimpl}})\\[1.25em]~\end{array}&
\semdefff
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\hspace{-1em}\begin{array}{l}  
      \coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} }\\[-0.3em]
        \quad \vdash \kvd{let}~\ident{?p}=e_1\\[-0.3em]
        \quad\qquad \kvd{in}~e_2 : \tau_2 \\[-0.3em]
        ~ \\[-0.3em]
   \end{array}}
  {f}
  {g}
  {\hspace{-1em}\begin{array}{l}  
  \lambda\ctx.~~\\[-0.3em]
    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cclrd{\cup}\; \cclrd{(s\setminus\{\ident{?p}\})}}~\ident{dup}~\ctx\\[-0.3em]
    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{(s\setminus\{\ident{?p}\})}}~\ctx_0\\[-0.3em]
    \quad g~(\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{(s\setminus\{\ident{?p}\})}}~(f~\ctx_1)~\ctx_2)
  \end{array} }   
\end{array}
\end{equation*}

\vspace{1em}
\paragraph{Reduction rules}

\begin{equation*}
\begin{array}{rl}
{\footnotesize(\emph{counit})} & \begin{array}{l}
\ident{counit}_{\cclrd{ \emptyset }}~(\ident{Impl} (v, \ldots))   \rightarrow v
\end{array}
\\[0.5em]
{\footnotesize(\emph{cobind})} & \begin{array}{l}
\ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})) \rightarrow
\\[-0.25em]
\quad \ident{Impl}(f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \} )
\end{array}
\\[1em]
{\footnotesize(\emph{merge})} & \begin{array}{l}
\ident{merge}_{\cclrd{r}, \cclrd{s}} 
              (~\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})  \\[-0.25em]
\hspace{4.4em}    \ident{Impl}(v', \{ \ident{?p}'_1 \mapsto v'_1, \ldots, \ident{?p}'_{n}\mapsto v'_n \})~) \rightarrow \\[-0.25em]
\quad \ident{Impl}((v, v'), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}\in\cclrd{r \setminus s}  \} \cup \{ \ident{?p}'_i \mapsto v'_i \;|\; \ident{?p}'\in\cclrd{s}  \})
\end{array}
\\[1.75em]
{\footnotesize(\emph{split})} & \begin{array}{l}
\ident{split}_{\cclrd{r}, \cclrd{s}} (\ident{Impl}((v,v'), \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
\\[-0.25em]
\quad \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} ), \ident{Impl}(v', \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \} )
\end{array}
\\[1em]
{\footnotesize(\emph{letimpl})} & \begin{array}{l}
\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v'~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
\\[-0.25em]
\quad \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i\;|\; \ident{?p}_i \in \cclrd{r}, \ident{?p}_i \neq \ident{?p} \} \cup \{ \ident{?p} \mapsto v' \} )
\end{array}
\\[1em]
{\footnotesize(\emph{lookup})} & \begin{array}{l}
\ident{lookup}_{\cclrd{\ident{?p}_i}} (\ident{Impl}(v, \{\ident{?p}_i\mapsto v_i\}) \rightarrow v_i
\end{array}
\end{array}
\end{equation*}

\vspace{-0.5em}
\caption{Additional constructs embedding implicit parameters into the language}
\label{fig:semantics-ext-impl}
\end{figure}

\newpage
~
\newpage
\paragraph{Properties.}

The core functional language subset (Figure~\ref{fig:semantics-target}) with the comonadically-inspired 
extensions (Figure~\ref{fig:semantics-ext}) and the domain-specific extensions for implicit 
parameters defined in Figure~\ref{fig:semantics-ext-impl} is safe. Furthermore, the 
translation from coeffect calculus produces well-typed programs. Together, these two properties 
also guarantee safety of the source coeffect calculus, when used with implicit parameters.

As for dataflow computations, we prove safety by extending the preservation
(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog})
for the core functional subset of the language, but it is worth noting that the key parts of
the proofs are centered around the new reduction rules for comonadically-inspired primitives
and newly defined \ident{Impl} values. These do not interact with the rest of the language in 
any unexpected ways.

\begin{theorem}[Well-typedness of the implicit parameters translation]
\label{thm:semantics-impl-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
and Figure~\ref{fig:semantics-ext-impl} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation.

\vspace{0.5em}\noindent\hangindent=0.6cm 
Case (\emph{lookup}): The type of $\ctx$ has a coeffect  $\cclrd{ \{\ident{?p}\} }$
  which includes the parameter \ident{?p} as required in order to use the (\emph{lookup}) typing rule.

\vspace{0.5em}\noindent\hangindent=0.6cm 
Case (\emph{letimpl}): The type of $\ctx$ matches with the input type of
  $\ident{map}_{  \cclrd{r \cup (s\setminus\{\ident{?p}\})} }$. After duplication and splitting
  the context, $\ctx_1$ and $\ctx_2$ have types $\ctyp{\cclrd{r}}(\ldots)$ and
  $\ctyp{\cclrd{s\setminus\{\ident{?p}\}}}(\ldots)$, respectively. This matches with the expected
  types of $\ident{f}$ and $\ident{letimpl}$. The context returned by $\ident{letimpl}$ then 
  matches the one required by $\ident{g}$.
\end{proof}

\begin{lemma}[Canonical forms]
\label{thm:semantics-impl-canon}
For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
\begin{enumerate}
  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
  \item If $\tau=\ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}\tau_1$ then $e = \ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_n \mapsto v_n \} )$
\end{enumerate}  
\end{lemma}
\begin{proof}
  (1,2,3) as before; for (4) the last typing rule must have been (\emph{impl}).
\end{proof}

\begin{lemma}[Preservation under substitution]
\label{thm:semantics-impl-pres-subst}
For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$ 
then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$ as before, with new
  cases for $\ident{Impl}(e, \{ \ldots \})$, $\ident{lookup}_{\cclrd{\ident{?p}}}$ and 
  $\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}$.
\end{proof}

\begin{theorem}[Type preservation]
\label{thm:semantics-impl-pres}
  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
\end{theorem}
\begin{proof}
  Rule induction over $\rightarrow$.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{fn}, \emph{prj}, \emph{ctx}): As before, using Lemma~\ref{thm:semantics-impl-pres-subst} for (\emph{fn}).

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{counit}): $e = \ident{counit}_{\cclrd{0}}(\ident{Impl}(v, \{\;\}))$. The last rule in the type
  derivation of $e$ must have been (\emph{counit}) with $\Gamma \vdash \ident{Impl}(v, \{ \; \}) : \ctyp{\cclrd{\emptyset}}\tau$
  which, in turn, required that $\Gamma \vdash v : \tau$.

@@

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n  \}))$.  
  The last rule in the type derivation of $e$ must have been (\emph{cobind}) with a type $\tau = \ctyp{\cclrd{s}}\tau_2$ and assumptions
  $\Gamma \vdash \ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n  \}) : \ctyp{\cclrd{r}}{\tau}$ and
  $\Gamma \vdash f : \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$.
  The reduced expression has a type $\ctyp{\cclrd{s}}\tau_2$:
\begin{equation*}
\inference
  { \inference 
      { \Gamma \vdash f : \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 & 
        \Gamma \vdash \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} ) :  \ctyp{\cclrd{r}}{\tau_1}}  
      { \Gamma \vdash f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )) : \tau_2} }
  { \Gamma \vdash \ident{Impl}(f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \}) : \ctyp{\cclrd{s}}\tau_2 }
\end{equation*}

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{lookup}): $e = \ident{lookup}_{\cclrd{\ident{?p}_i}}(\ident{Impl}(v, \{\; \ldots, \ident{?p}_i \mapsto v_i \ldots \}))$. The last rule in the type
  derivation must have been (\emph{lookup}) with $\tau = \ident{num}$ and an assumption $\Gamma\vdash\ident{Impl}(v, \{\; \ldots, \ident{?p}_i\mapsto v_i \ldots \}) : \ctyp{\cclrd{ \{\ldots,\ident{?p}_i,\ldots\} }}{\tau}$,
  which requires $\Gamma \vdash v_i : \ident{num}$.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{merge}, \emph{split}, \emph{letimpl}): Similar. In all three cases, the last typing rule in the derivation of $e$
  guarantees that all values of all implicit parameters that are required for the reduction are available.
\end{proof}

\begin{theorem}[Progress]
\label{thm:semantics-impl-prog}  
  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
\end{theorem}
\begin{proof}
  By rule induction over $\vdash$.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{num,abs,var,app,proj,tup}): As before, using the adapted canonical forms lemma
  (Lemma~\ref{thm:semantics-impl-canon}) for (\emph{app}) and (\emph{proj}).

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{counit}): $e = {\ident{counit}_\cunit}~e_1$. If $e_1$ is not a value, it can be reduced 
  using (\emph{ctx}) with context ${\ident{counit}_\cunit}~\_$, otherwise it is a value. From Lemma~\ref{thm:semantics-df-canon},
  $e_1 = \ident{Impl}(v, \{\;\})$ and so we can apply (\emph{counit}) reduction rule.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2$. If $e_1$ is not a value,
  reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{r},\cclrd{s}}}~\_~e$. If $e_2$ is
  not a value reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{r},\cclrd{s}}}~v~\_$.
  If both are values, then from Lemma~\ref{thm:semantics-impl-canon}, we have 
  $e_2 = \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in \cclrd{r}\cup\cclrd{s} \})$ 
  and we apply the (\emph{cobind}) reduction.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{merge}): $e = \ident{merge}_{\cclrd{r}, \cclrd{s}} e_1$. If $e_1$ is not a value, 
  reduce using (\emph{ctx}) with context $e = \ident{merge}_{\cclrd{r}, \cclrd{s}}~\_$. If $e_1$ is
  a value, it must be a pair of values $(   
    \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in\cclrd{r} \}),
    \ident{Impl}(v', \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in\cclrd{s} \}))$
  using Lemma~\ref{thm:semantics-df-canon} and it can reduce using (\emph{merge}) reduction.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{impl}): $e = \ident{Impl}(e', \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$. 
  If $e$ is not a value, reduce using (\emph{ctx}) with context
  $\ident{Impl}(\_, \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$. 
  If $e_i$ is not a value, reduce using (\emph{ctx}) with context 
  $\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{i-1} \mapsto v_{i-1}, \ident{?p}_i \mapsto \_, \ident{?p}_{i+1} \mapsto v_{i+1}, \ldots \ident{?p}_n \mapsto e_n\})$.
  Otherwise, $e, e_0, \ldots, e_n$ are values and so 
  $\ident{Impl}(e', \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$
  is also a value.

\vspace{0.25em}\noindent\hangindent=0.6cm 
Case (\emph{split}, \emph{letimpl}): Similar. Either sub-expression is not a value, 
  or the type guarantees that it is a comonadic value with implicit parameter bindings
  that enable the (\emph{split}) or (\emph{letimpl}) reduction, respectively.
\end{proof}
